SPANG: Template Library for Making SPARQL Queries FAIR

### 2020-05-24 paper.tex
v2.1

An increasing number of datasets have been published in the Resource Description Framework (RDF)~\cite{rdf}.
SPARQL is the standardized interface to RDF and contributing effective reuse of distributed RDF data. 
How can we maximize the value of accumulated SPARQL~\cite{sparql} queries?
Reuse of the queries is the key of this problem.
Although eco-systems for developing SPARQL have been developed, writing SPARQL codes often becomes a burden for the Semantic Web developers and users. 
SPARQL queries have been written in many Semantic Web projects, which requires efforts. 
For effective construction of the Semantic Web applications, reuse of such queries are valuable.
% Reusability is the key to maximize the value of accumulated open data and software.
However, due to the limitation of SPARQL specifications, such as the lack of parameterization mechanism, the reuse of those queries are limited. 
% Reusing SPARQL queries are important for efficient development of the Semantic Web.

%The FAIR principle~\cite{fair} could be suitable for addressing this issue.
One of the key principle in recent days of the Semantic Web community is the FAIR (findable, accessible, interoperable and reusable) principle~\cite{fair}.
The ultimate goal of FAIR is to optimise the reuse of resources. 
The Semantic Web is comprised of various specifications, including RDF~\cite{rdf}, OWL~\cite{owl}, and SPARQL~\cite{sparql}. The ontologies written in OWL are accumulated and shared on repositories~\cite{bioportal}.
RDF datasets are also accumulated and hosted in portal web sites~\cite{rdf-portal}.
Not only the ontologies and datasets, but also the queries are important for the Semantic Web.
% As the SPARQL queries are important component of the Semantic Web community,
% they should be shared in a solid principle.
Accumulating and reusing the queries are important.
Notably, the FAIR principle can be applied not only to the RDF data, but also other resource, like software.
% SPARQL is a fundamental technology for querying RDF data contributing an interoperable interface to the Semantic Web. However, is is not designed for reusability. 
% In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources. 
Applying FAIR principle to SPARQL query should increase the reusability of the queries.

@@

Here, we apply the FAIR principle and extend the SPARQL specifications for more efficient development of the Semantic Web.
%Semantic Web users often write SPARQL queris.
%However, the SPARQL lacks the parameterization mechanism, thus, users write queries for their specific use cases.
%But good SPARQL queries can be reused.
%Actually, writing good query requires much efforts, thus, they should be reused to improve the productivity.
%Thus, considering the reusability of SPARQL queries are crucial to enrich the Semantic Web community.
%How can we improve the reusability of SPARQL?
Following the FAIR principle, we could identify the process to make SPARQL query reusable.
One of the key to improve the reusability of queries is parameterization.
Parameterization make the query more reusable.
Using the template query with parameters, the same query can be applied to many use cases.
% There can be several types of parameterization.
% The SPARQL variables can be considered as parameter.
% In that case, values to the parameters can replace them.
% On the other hands, part of the URI or literals can be parameterized.
% In that case also, values to the parameters can replace them.
% Another mechanism to use template is to give binding to the variables.
%Are SPARQL queries accessible? 
Accessibility to SPARQL queries should also be reconsidered.
In most cases, those queries are not accessible. According to the FAIR principle, resources should be accessible by through URI. In some cases, SPARQL queries are stored in repositories. But they do not have stable URIs. How about findability? 
In many Semantic Web projects, 
%SPARQL queries are written and accumulated. However, those SPARQL queries 
%are not findable, due to the lack of export and import SPARQL queries.
SPARQL queries are not exposed to the web in a reusable form, nor indexed by metadata in a findable form.

In this study, we show the steps to make SPARQL queries FAIR to maximize the value of written SPARQL queries:
1) assign meta data to SPARQL queries;
2) parameterize the SPARQL queries;
3) assign global identifiers;
4) construct a server for providing queries.
And we demonstrate the steps to construct the SPANG template library, which is reusable on the Semantic Web platform.

@@

Figure~\ref{fig:spang_top} is the top page of SPANG server, which provides access to a SPANG library, links to GitHub repositories of the SPANG project, and documentations. The SPANG project consists of SPANG server, SPANG library, SPANG API, and SPANG command. The central topic of this paper is how to construct the SPANG library in an interoperable and reusable way. It also provides the foundation of findability and accessibility through SPANG API and command.

In the following sections, we review FAIR principles and describe how to apply the principle to SPARQL.
Further, we demonstrate how to use the SPANG template library.
Then, we describe the details of SPANG template specifications.

@@

\begin{figure}
\center
\includegraphics[width=1.0\textwidth]{spang_top.png}
\caption{SPANG server top page}
\label{fig:spang_top}
\end{figure}

@@

The FAIR guiding principle aims to enhance the Findability, Accessibility, Interoperability and Reusability of digital resources.
It shoul be noted that the FAIR principle can be applied not only to the dataset resources but also to other type of resources including software tools.
Not only the RDF dataset, but also the SPARQL queries are important resource to be shared on the Semantic Web platform. Thus, we tried to apply the FAIR principle which have been applied to the RDF datasets, into SPARQL queries.
In the following subsections, we review the preliminaries for the FAIR principle
and provide the FAIR principle in the context of SPARQL queries.
We review each item in the following subsections.
In the following subsections, the target of FAIR principle is extended from data to resource.
We consider a SPARQL query as a resource.

@@

\subsection{Findability}
The ultimate goal of FAIR is to optimise the reuse of resources. 
To achieve this, metadata should be well-described so that they can be replicated and/or combined in different settings.
In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
Most of the SPARQL queries are not findable from general users of the Semantic Web. 
The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
The first step is using metadata to find SPARQL queries. 
Metadata should be easy to find for both humans and computers. Machine-readable metadata are essential for automatic discovery of resources, so this is essential component.
% \begin{itemize}
%     \item Assign global identifier
%     \item Rich metadata
%     \item Include the identifier of the data they describe
%     \item Registered and indexed
% \end{itemize}
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes: title, description, target endpoint, and parameters with its name, default value, and classes.
Indexing of SPARQL queries are conducted and thus users can obtain specific queries of interest through application programming interfaces (APIs).

@@

\subsection{Interoperability}
For interoperability, a formal, accessible, shared, and broadly applicable language is necessary.
The resource usually need to be integrated with other resources. 
In addition, the resource need to interoperate with application or workflows for analysis, storage, and processing.
% \begin{itemize}
%     \item Use a formal, accessible, shared, and broadly applicable language for knowledge representation.
%     \item Use vocabularies that follow FAIR principles.
%     \item Include qualified references to others
% \end{itemize}

SPARQL specification is proposed for interoperable use of queries to the Semantic Web. 
It is valuable to use several triplestore implementation through the same programming interface. 
However, standard SPARQL specifications do not include coding rules; capitalization, indentation and newline, etc. 
Lack of those rules result in variation of SPARQL queries, which hampers the efficiency in collaborative development of SPARQL queries. 
In other languages, coding rules exist in most cases. 
In the case of SPARQL, even specification document or text books do not follow a specific coding rules. 
Thus, we specify the coding rule used in SPANG template library, and implemented a formatter command to evaluate and reformat the input SPARQL code.

@@

%Reusability of SPARQL is not a new topic. 
The ultimate goal of FAIR is to optimise the reuse of resources. 
However, in the specification of SPARQL, reusability is not a central topic. 
Since it is an central resource in a Semantic Web, it is valuable to share and reuse SPARQL queries.
Templating is a key for reusability. 
SPARQL specifications do not includes templating mechanism, but without templating, each SPARQL queries are not applicable to other use cases. 
We implemented templating mechanism using a general freamework, mustache.
Various activity is conducted to construct libraries of SPARQL. 
It is considered as activities to increase the reusability of SPARQL queries. However, no standard to increase the reusability is proposed. 
%If it is a component of Semantic Web comprised of RDF/OWL/SPARQL, it is reasonable to represent it in RDF framework. 
%We propose a way to represent SPARQL in the RDF framework.

@@

\end{verbatim}
\end{scriptsize}
\caption{Example query in the SPANG library}
\label{fig:example-sparql}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SPANG Library and Use Cases}
%%%%%%%%%%%%%%%%%%%%%%%%%
In the previous section we review the FAIR principle in the context of SPARQL queries.
In this section, we describe how we constructed the SPANG library with some use cases.
The use case shown in this section includes the query to the DisGeNet endpoint~\cite{disgenet}, which is shown in Figure~\ref{fig:example-sparql}.
This query retrieve protein IDs involved in a specific disease.
The core of the SPANG library specification is the SPARQL metadata and templating mechanism for increasing the reusability of the SPARQL queries.
The query example includes metadata at the beginning of the query as comment lines.
The query also includes a parameter for a disease identifier.
According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.

\subsection{SPARQL Metadata}

We propose a set of SPARQL metadata to annotate each SPARQL query, to make them more reusable.
Those metadata are added as comment at the head of the SPARQL query, thus that does not violate the SPARQL specification. 
The set of metadata increase the findability of SPARQL queries.

@@

Figure~\ref{fig:spang_lib} is the list of SPARQL templates for DisGeNET SPARQL endpoint.
A short name for the SPARQL template, description of the query, and example parameters are shown.

@@

\begin{figure}
\center
\includegraphics[width=1.0\textwidth]{spang_disgenet.png}
\caption{SPANG templates for DisGeNet SPARQL endpoint}
\label{fig:spang_disgenet}
\end{figure}

@@

SPANG is implemented in JavaScript using the Node.js technology~\cite{nodejs}. Thus, it can be called in JavaScript applications, and also in the command-line environment.

\subsection{Template Specifications}
SPANG template syntax is an extension of sparql-doc. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a syntax parameterization, we consider mustache~\cite{mustache} notation.

\subsection{Template Engine}
SPANG template engine is implemented, where each template can be transformed into abstract syntax tree, and then reconstructed into SPARQL. 
We used PEG~\cite{peg} expression of SPARQL grammar originally expressed in EBNF in the SPARQL specification. The original implementation of PEG.js~\cite{pegjs} of SPARQL grammar is provided by rdfstore-js project (https://github.com/antoniogarrote/rdfstore-js).
The PEG.js code was modified to fit the use in the case of SPANG library.
The PEG.js is a parser generator, thus, we could generate our own parser based on the PEG.js description. Thus, we obtained abstract syntax tree in the form of JSON object.

@@

The template can be URL that returns a SPARQL query, such as \url{https://cdn.jsdelivr.net/gh/hchiba1/spang-library@latest/library/disgenet/disease_gene.rq}. The full usages of the command is as shown in Figure~\ref{fig:spang-command}.

@@

\end{verbatim}
\end{scriptsize}
\caption{Usage of the \texttt{SPANG2} command}
\label{fig:spang-command}
\end{figure}

>>>>

An increasing number of datasets have been published in the Resource Description Framework (RDF)~\cite{rdf}.
SPARQL~\cite{sparql} is the standardized interface to RDF and contributing effective reuse of distributed RDF data. 
Although eco-systems for developing SPARQL have been developed, writing SPARQL codes often becomes a burden for the Semantic Web developers and users. 
SPARQL queries have been written in many Semantic Web projects, which are the result of developers' efforts. 
How can we maximize the value of accumulated SPARQL queries?
Reuse of the queries is the key of this problem.
For effective construction of the Semantic Web applications, reuse of such queries are valuable.
% Reusability is the key to maximize the value of accumulated open data and software.
However, due to the limitation of SPARQL specifications, such as the lack of parameterization mechanism, the reuse of those queries are limited. 
% Reusing SPARQL queries are important for efficient development of the Semantic Web.

%The FAIR principle~\cite{fair} could be suitable for addressing this issue.
One of the key principle in recent days of the Semantic Web community is the FAIR (findable, accessible, interoperable and reusable) principle~\cite{fair}.
The ultimate goal of FAIR is to optimise the reuse of resources. 
The Semantic Web is comprised of various specifications, including RDF~\cite{rdf}, OWL~\cite{owl}, and SPARQL~\cite{sparql}. 
RDF datasets are also accumulated and hosted in portal web sites~\cite{rdf-portal}.
The ontologies written in OWL are accumulated and shared on repositories~\cite{bioportal}.
Not only the datasets and ontologies, but also the queries are important for the Semantic Web.
% As the SPARQL queries are important component of the Semantic Web community,
% they should be shared in a solid principle.
Accumulating and reusing the queries are important.
Notably, the FAIR principle can be applied not only to the RDF data, but also other resource, like software.
% SPARQL is a fundamental technology for querying RDF data contributing an interoperable interface to the Semantic Web. However, is is not designed for reusability. 
% In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources. 
Thsu, applying FAIR principle to SPARQL query should increase the reusability of the queries.

@@

Here, we apply the FAIR principle and extend the SPARQL specifications for more efficient development of the Semantic Web applications.
%Semantic Web users often write SPARQL queris.
%However, the SPARQL lacks the parameterization mechanism, thus, users write queries for their specific use cases.
%But good SPARQL queries can be reused.
%Actually, writing good query requires much efforts, thus, they should be reused to improve the productivity.
%Thus, considering the reusability of SPARQL queries are crucial to enrich the Semantic Web community.
%How can we improve the reusability of SPARQL?
According to the FAIR principle, we identified the process to make SPARQL query reusable.
One of the key to improve the reusability of queries is parameterization.
Parameterization make the query more reusable.
Using the template query with parameters, the same query can be applied to many use cases.
% There can be several types of parameterization.
% The SPARQL variables can be considered as parameter.
% In that case, values to the parameters can replace them.
% On the other hands, part of the URI or literals can be parameterized.
% In that case also, values to the parameters can replace them.
% Another mechanism to use template is to give binding to the variables.
%Are SPARQL queries accessible? 
Accessibility to SPARQL queries should also be reconsidered.
In most cases, those queries are not accessible. According to the FAIR principle, resources should be accessible by through URI. In some cases, SPARQL queries are stored in repositories. But they do not have stable URIs. How about findability? 
In many Semantic Web projects, 
%SPARQL queries are written and accumulated. However, those SPARQL queries 
%are not findable, due to the lack of export and import SPARQL queries.
SPARQL queries are not exposed to the web in a reusable form, nor indexed by metadata in a findable form.

In this study, we show the steps to make SPARQL queries FAIR to maximize the value of written SPARQL queries:
1) assign meta data to SPARQL queries;
2) parameterize the SPARQL queries;
3) assign global identifiers;
4) construct a server for providing queries.
Further, we demonstrate the steps to construct the SPANG template library, which is reusable on the Semantic Web platform.

@@

The SPANG~\cite{spang} project stared aiming at maximizing the usability of SPARQL queries, and the project now includes development of SPANG server, SPANG library, SPANG API, and SPANG command. 
Figure~\ref{fig:spang_top} is the top page of SPANG server, which provides access to a SPANG library, links to GitHub repositories of the SPANG project, and documentations. 
The central topic of this paper is how to construct the SPANG library in an interoperable and reusable way. 
The SPANG library specifications also provide the foundation of findability and accessibility through SPANG API and command.

In the following sections, we review FAIR principles in the context of SPARQL and describe how to apply the principle to SPARQL.
Further, we demonstrate how to use the SPANG template library.
Then, we describe the details of SPANG template specifications.

@@

\begin{figure}
\center
\includegraphics[width=1.0\textwidth]{spang_top.png}
\caption{SPANG project top page}
\label{fig:spang_top}
\end{figure}

@@

The FAIR guiding principle~\cite{fair} aims to enhance the Findability, Accessibility, Interoperability and Reusability of digital resources.
It should be noted that the FAIR principle can be applied not only to datasets but also to other type of resources including software.
Not only the RDF dataset, but also the SPARQL queries are important resource to be shared on the Semantic Web platform. Thus, we tried to apply the FAIR principle which have been applied to the RDF datasets, into SPARQL queries.
In the following subsections, we review the preliminaries for the FAIR principle
and provide the FAIR principle in the context of SPARQL queries.
We consider a SPARQL query as a resource,
and here, the target of FAIR principle is extended from data to resources which include SPARQL queries.

@@

\subsection{Findability}
%The ultimate goal of FAIR is to optimise the reuse of resources. 
%To achieve this, metadata should be well-described so that they can be replicated and/or combined in different settings.
In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
Most of the SPARQL queries are not findable from general users of the Semantic Web. 
The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
The first step is using metadata to find SPARQL queries. 
Metadata should be easy to find for both humans and computers. Machine-readable metadata are essential for automatic discovery of resources, so this is essential component.
% \begin{itemize}
%     \item Assign global identifier
%     \item Rich metadata
%     \item Include the identifier of the data they describe
%     \item Registered and indexed
% \end{itemize}
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes: title, description, target endpoint, and parameters with its name, default value, and classes.
Indexing of SPARQL queries are conducted and thus users can obtain specific queries of interest through application programming interfaces (APIs).

@@

\subsection{Interoperability}
%For interoperability, a formal, accessible, shared, and broadly applicable language is necessary.
For interoperability, a shared language is necessary.
The resource usually need to be integrated with other resources. 
In addition, the resource need to interoperate with application or workflows for analysis, storage, and processing.
% \begin{itemize}
%     \item Use a formal, accessible, shared, and broadly applicable language for knowledge representation.
%     \item Use vocabularies that follow FAIR principles.
%     \item Include qualified references to others
% \end{itemize}

SPARQL specification is proposed for interoperable use of queries to the Semantic Web. 
Indeed, it is valuable to use several triplestore implementation through the same programming interface. 
Nevertheless, standard SPARQL specifications do not include coding rules; capitalization, indentation and newline, etc. 
Lack of such rules result in variation of SPARQL queries, which hampers the efficiency in collaborative development of SPARQL queries. 
In the programming languages in general, coding rules exist in most cases. 
In the case of SPARQL, however, even specification document or text books do not follow a specific coding rules. 
Thus, we specify the coding rule used in SPANG template library, and implemented a formatter command to evaluate and reformat the input SPARQL.

@@

%Reusability of SPARQL is not a new topic. 
The ultimate goal of FAIR is to optimise the reuse of resources. 
However, in the specification of SPARQL, reusability is not considered as a central topic. 
Since SPARQL queries is essential resources comprising the Semantic Web, it is valuable to share and reuse SPARQL queries.
Templating is a key for improving reusability. 
SPARQL specifications do not includes templating mechanism, but without templating, each SPARQL queries are not applicable to other use cases. 
We implemented templating mechanism using a general freamework, mustache~\cite{mustache}.
Various activities are conducted to construct libraries of SPARQL~\cite{bioqueries}. 
It is considered as activities to increase the reusability of SPARQL queries. However, no standard to increase the reusability is proposed. 
%If it is a component of Semantic Web comprised of RDF/OWL/SPARQL, it is reasonable to represent it in RDF framework. 
%We propose a way to represent SPARQL in the RDF framework.

@@

\end{verbatim}
\end{scriptsize}
\caption{Example query in the SPANG template library}
\label{fig:example-sparql}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SPANG Template Library with Use Cases}
%%%%%%%%%%%%%%%%%%%%%%%%%

In the previous section we review the FAIR principle in the context of SPARQL queries.
In this section, we describe how we constructed the SPANG library with some use cases.
The core of the SPANG library specification is the SPARQL metadata and templating mechanism for increasing the reusability of the SPARQL queries.
According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.
The use case shown in this section includes the query to the DisGeNET endpoint~\cite{disgenet} shown in Figure~\ref{fig:example-sparql}.
This query retrieves protein IDs involved in a specific disease.
The query example includes metadata at the beginning of the query as comment lines.
The query also includes a parameter for a disease identifier.

Figure~\ref{fig:spang_disgenet} is the list of SPARQL templates for DisGeNET SPARQL endpoint.
A short name for the SPARQL template, description of the query, and example parameters are shown.

\begin{figure}
\center
\includegraphics[width=1.0\textwidth]{spang_disgenet.png}
\caption{SPANG templates for DisGeNET SPARQL endpoint}
\label{fig:spang_disgenet}
\end{figure}

\subsection{Assigning Metadata}

We propose a set of SPARQL metadata to annotate each SPARQL query, to make them more findable.
Those metadata are added as comment at the head of the SPARQL query, thus that does not violate the SPARQL specification. 
The set of metadata increase the findability of SPARQL queries.

@@

@@

@@

SPANG is implemented in JavaScript using the Node.js~\cite{nodejs}. Thus, it can be called in JavaScript applications, and also in the command-line environment.

\subsection{Template Specifications}
SPANG template syntax is an extension of sparql-doc~\cite{sparql-doc}. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a syntax parameterization, we consider mustache~\cite{mustache} notation.

\subsection{Template Engine}
SPANG template engine is implemented, where each template can be transformed into abstract syntax tree, and then reconstructed into SPARQL. 
We used Parsing Expression Grammar (PEG)~\cite{peg} expression of SPARQL grammar originally expressed in EBNF in the SPARQL specifications~\cite{sparql}. The original implementation of PEG.js~\cite{pegjs} of SPARQL grammar is provided by rdfstore-js project (https://github.com/antoniogarrote/rdfstore-js).
The PEG.js code was modified to fit the use in the case of SPANG library.
The PEG.js is a parser generator, thus, we could generate our own parser based on the PEG.js description. Thus, we obtained abstract syntax tree in the form of JSON object.

@@

The template can be URL that returns a SPARQL query, such as \url{https://cdn.jsdelivr.net/gh/hchiba1/spang-library@latest/library/disgenet/disease_gene.rq}. The full usages of the SPANG command is as shown in Figure~\ref{fig:spang-command}.

@@

\end{verbatim}
\end{scriptsize}
\caption{Usage of the SPANG command}
\label{fig:spang-command}
\end{figure}
